---
title: "Capstone - Movie Ratings Prediction"
output: 
  html_notebook:
    fig_caption: true
    toc: true
    toc_float: true
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE)
```

# Project Overview
This is a capstone project for [HarvardX Data Science - Professional Certificate](http://rmarkdown.rstudio.com) program. 
Objective of project is to create a move recommendation system using MovieLens dataset, using all tools (in particular ML algorithms) shown throught-out the HarvardX DS program.

## Dataset
Data assigned for this project is the 10 million ratings version of the [MovieLens dataset](https://grouplens.org/datasets/movielens/10m/).
There are 72K users in this dataset.
The dataset is created as follows:

```{r train_validation_dataset}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
install.packages("kableExtra")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(stringr)
library(knitr)
library(kableExtra)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")


# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

#add features to original dataset, so test set has the same features
#features- release year, title, rating date, rating hour
movielens <- movielens %>%
  mutate(release_yr = parse_number(str_split(title, pattern = "\\(", simplify = TRUE)[,2])) %>%
  mutate(title = str_split(title, pattern = "\\(", simplify = TRUE)[,1]) %>%
  mutate(rating_date = date(as_datetime(timestamp[1]))) %>%
  mutate(rating_hour = hour(as_datetime(edx$timestamp[1])))

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

## Data familiarization

```{r dim}
rows <- dim(edx)[1]
cols <- dim(edx)[2]
```
The `edx` training dataset has `r rows` rows (number of ratings) and  `r cols` columns.  

### Exploratory Data Analysis
#### Rating Distribution

Frequency distribution of ratings are shown in table 1 and figure 1 below:
```{r rating_freq_tab}
rft <- descr::freq(as.factor(edx$rating), plot = FALSE)
zero_rating <- edx %>% filter(rating == 0) %>%
  nrow()
three_rating <- edx %>% filter(rating == 3) %>%
  nrow()
kable(rft)
```
No. of ratings given zero = `r zero_rating`.
No. of ratings given three = `r three_rating`.

```{r rating_hist, fig.cap = "Fig. 1.1 Freq. dist. of ratings."}
library(hrbrthemes)
library(scales)
x <- seq(0, 5, 0.5)
edx %>% ggplot(aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  stat_bin(binwidth = 0.5, geom ="text", aes(label = ..count..), vjust = -1.5) +
  theme_ipsum() +
  scale_y_continuous(label = label_number_si(), limits = c(0, 3e6)) +
  scale_x_continuous(labels = as.character(x) ,breaks = x)
```

No. of distinct movies rated = `r n_movies`.
```{r n_distinct_movies}
n_movies <- n_distinct(edx$movieId)
n_movies
```
No. of distinct movies by rating
```{r movie_dist_rating, fig.cap = "Fig 1.2 # of unique movies by ratings"}
edx %>% group_by(rating) %>% summarize(n_mov = n_distinct(movieId)) %>%
  ggplot(aes(x = rating, y = n_mov)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label= n_mov, vjust = -1.5)) +
  ylim(0, 12500) +
  scale_x_continuous(labels = as.character(x) ,breaks = x) +
  ylab("no of movies") +
  theme_ipsum()
```

No. of unique users = `r n_users`.
```{r n_unique_users}
n_users <- n_distinct(edx$userId)
n_users
```

No. of distinct users by rating
```{r user_dist_rating, fig.cap = "Fig 1.3 # of unique users by ratings"}
edx %>% group_by(rating) %>% summarize(n_user = n_distinct(userId)) %>%
  ggplot(aes(x = rating, y = n_user)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label= n_user, vjust = -1.5)) +
  scale_y_continuous(labels = label_number_si(), limits = c(0, 80000)) +
  scale_x_continuous(labels = as.character(x) ,breaks = x) +
  ylab("no of users") +
  theme_ipsum()
```

Genre stats
No. of unique genre combinations
```{r n_distinct_genre_combo}
n_genres_combo <- n_distinct(edx$genres)
n_genres_combo
```

Genres that are available in the dataset include:
```{r distinct_genres}
n_drama <- edx %>% filter(str_detect(edx$genres, "Drama") == TRUE) %>% summarize(n = n())

starttime <- Sys.time()

genre_tab <- edx[,c(1:2,6)] %>% separate_rows(genres, sep = "\\|") %>%
  group_by(genres) %>%
  summarize(n_movies = n_distinct(movieId), n_ratings = n()) 
endtime <- Sys.time()

duration <- endtime - starttime
duration #18.7mins 

kable(genre_tab)
```

No. of movie ratings for Drama genre is `r n_drama`.

```{r genre_mov_dist, fig.cap = "Fig 1.3 No. of movies by Genres" }
p <- genre_tab %>% ggplot(aes(x = reorder(genres, -n_movies), y = n_movies)) +
  geom_bar(stat = "identity")
p <- p +
  labs(x = "genres",
       y = "# of movies") +
  coord_flip() +
  theme_classic() #most common genre is Drama
p + geom_text(aes(label = n_movies), vjust = 0, hjust = -0.5) +
  ylim(c(0, 6000))
```
```{r genre_combo_mov_dist, fig.cap = "Fig 1.4 Most Popular Genre Combinations"}
p2 <- edx %>% group_by(genres) %>% 
  summarize(n_movies = n_distinct(movieId)) %>%
  filter(n_movies > 100) %>%
  ggplot(aes(x = reorder(genres, -n_movies), y = n_movies)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_movies), hjust = -0.5) +
  ylim(c(0,2000))
p2 <- p2 + coord_flip() +
  theme_classic() +
  labs(y = "# of movies",
       x = "genres")
p2
```

Indicates in terms of genres, there is a higher proportion of movies in the database
that are combination of drama, comedy romance. Followed by documentary & horror.

# lets build a table of frequencies by genre 20X2 and 797x2 tables that is sorted


#model answer
# str_detect
genres = c("Drama", "Comedy", "Thriller", "Romance")
sapply(genres, function(g) {
  sum(str_detect(edx$genres, g))
})

# separate_rows, much slower!
# edx %>% separate_rows(genres, sep = "\\|") %>%
  #group_by(genres) %>%
  #summarize(count = n()) %>%
  #arrange(desc(count))

title <- edx %>% group_by(title) %>% summarise(n = n()) %>% arrange(desc(n))

edx %>% group_by(movieId, title) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

edx %>% group_by(rating) %>% summarize(count = n()) %>% arrange(desc(count))

#model answer
edx %>% group_by(rating) %>% summarize(count = n()) %>% top_n(5) %>%
  arrange(desc(count))

check <- edx %>% mutate(status = if_else(rating %% 1 == 0, "whole", "half")) %>%
                 group_by(status) %>% summarize(count = n())
#model answer
edx %>%
  group_by(rating) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = rating, y = count)) +
  geom_line()

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
